---
title: "Chapter 11 Practice"
author: "Tammy Lightle"
date: today
format: html
editor: visual
---

## Task 0: Load packages

```{r}
#| warning: false

library(dslabs)
library(dplyr)
```

## Task 1: Column Selection

```{r}
C_Overload <- select(temp_carbon, year, carbon_emissions)

head(C_Overload)
```

Select kept the two columns: year and carbon_emissions. Head verified this by showing the top 6 rows.

## Task 2: Single-Column Dataframe

```{r}
star_names <- stars %>%
  select(star)

head(star_names)
```

This returns a one-column data frame, with the column `star`.

## Task 3: Row Filtering

```{r}
unique(murders$state)

No_Florida <- murders %>%
  filter(state != "Florida")

unique(No_Florida$state)
```

Pre-checking the values may be useful so that you know what states exist in the original dataset. In the verify step, I should not see Florida listed. All rows where Florida is the state are removed from the No_Florida data frame.

## Task 4: Multiple Conditions

```{r}
summary(mtcars$qsec)

Fast_n_Furious <- mtcars %>%
  filter(am == 1,
         cyl == 8,
         qsec < 15)

head(Fast_n_Furious)
nrow(Fast_n_Furious)
```

I filtered out all cars that meet three conditions. First, the car must have a manual transmission (`am == 1`). Second, the car must have eight cylinders (`cyl == 8`). Finally, the car must have a 1/4 mile time of less than 15 seconds (`qsec < 15`). Two cars meet all of these conditions.

## Task 5: MISSING

## Task 6: OR Logic + Column Slice

```{r}
range(mtcars$disp)

filtered_cars <- mtcars %>%
  filter(disp < 160 | disp > 250) %>%
  select(1:3)

all(filtered_cars$disp > 250 | filtered_cars$disp < 160)
```

There are 27 cars that have a displacement of less than 160 or greater than 250. Using `all()` shows that each row meets that criteria.

## Task 7: Reflection

I find that using the `dplyr` verbs of `select()` and `filter()` is more straightforward than the base-R brackets `[]` for subsetting. It is easier to read a long code section that uses the `%>%` operator!

## Task 8: Self-check checklist

Sanity check - passed!

##  
