---
title: "Chapter 16 Practice"
author: "Tammy Lightle"
date: today
toc: true
toc-depth: 2
number-sections: true
format: 
  html:
    code-fold: true
    self-contained: true
editor: visual
---

```{r setup}
#| warning: false

library(dslabs)
library(dplyr)
library(ggplot2)
```

## Task 1: Quick Recall

`ggplot2` has three core steps. First supply a dataframe, then map the aesthetics, and finally add geometric objects.

## Task 2: Scatterplot

Using the `iris` dataset, create a scatterplot showing the relationship between sepal length and petal length.

```{r}
Pretty_Flowers <- iris %>%
  ggplot(aes(Sepal.Length, Petal.Length, color = Species)) +
  geom_point(size = 3) +
  xlab("Sepal Length (cm)") +
  ylab("Petal Length (cm)")

Pretty_Flowers
```

The species virginica and versicolor have a stronger linear correlation between sepal length and petal length than setosa.

## Task 3: Boxplot - group comparison

```{r}
Petal_Spread <- iris %>%
  ggplot(aes(x = Species, y = Petal.Width, color = Species)) +
  geom_boxplot(show.legend = FALSE)

Petal_Spread
```

The species virginica has the largest petal width distribution.

## Task 4: Histogram vs Density Plot

```{r}
Tall_Density <- heights %>%
  ggplot(aes(x = height)) +
  geom_histogram(aes(y = after_stat(density)),
                     binwidth = 1,
                     fill = "blue",
                     color = "white") +
  geom_density(color = "darkorange",
               fill = "darkorange",
               alpha = 0.3,
               linewidth = 1) +
  xlab("Height (inches)") +
  ylab("Density")

Tall_Density
```

Combining the density curve with the histogram makes the shape easier to describe because it is easier to visualize the distribution.

## Task 5: Reflect

A density curve is more useful than a histogram when you want to describe the overall shape of the distribution. It will also be useful when comparing distributions between variables.
