---
title: "Chapter 13 Practice"
author: "Tammy Lightle"
date: today
toc: true
toc-depth: 2
number-sections: true
format: 
  html:
    code-fold: true
    self-contained: true
execute:
  warning: true
  message: true
editor: visual
---

```{r}
#| warning: false

library(dplyr)
library(dslabs)
```

## Task 1: Single Grouping

Compute the mean `temp` and mean `magnitude` for each star type in the `stars` dataset.

```{r}
twinkle_twinkle <- stars %>%
  group_by(type) %>%
  summarize(mean_temperature = mean(temp),
            mean_magnitude = mean(magnitude),
            .groups = "drop")

twinkle_twinkle
```

The result is a 10x3 tibble with columns `type`, `mean_temperature`, and `mean_magnitude`. There are 10 star types total. Star type O has the highest mean temperature, and star type M has the lowest. Star type DB has the highest mean magnitude, while star type O has the lowest.

## Task 2: Single Grouping + Three Stats

Find the region with the highest mean percent `stearic` acid of sample in the `olive` dataset.

```{r}

Olive_Garden <- olive %>%
  group_by(region) %>%
  summarize(mean_value = mean(stearic),
            sd_value = sd(stearic),
            n = n())

Olive_Garden
```

The highest mean percent `stearic` acid of sample was observed in Northern Italy. I wanted stats related to `stearic` acid, so that was the summary variable, and I wanted to view results by region, so `region` is the grouping variable.

## Task 3: Counting Unique Values

Determine the number of unique diseases in the `disease` column in the `us_contagious_diseases` dataset.

```{r}
length(unique(us_contagious_diseases$disease))
```

There are `{r} length(unique(us_contagious_diseases$disease))` unique diseases within the `disease` column in the `us_contagious_diseases` dataset. They are `{r} unique(us_contagious_diseases$disease)`.

Next, determine the number of states represented in the `state` column.

```{r}
length(unique(us_contagious_diseases$state))
```

There are `{r} length(unique(us_contagious_diseases$state))` unique states within the `state` column of the `us_contagious_diseases` dataset.

## Task 4: Make and Test a Prediction

If every state has 7 diseases, I would expect 357 (51 \* 7) rows when grouping by `state` and `disease` because each row should represent one state/disease combination.

Let's test:

```{r}
categorical_counts <- us_contagious_diseases %>%
  group_by(state, disease) %>%
  count()

nrow(categorical_counts)
```

My prediction was correct. It is useful to use `group_by()` to find out the frequency of multiple categorical data.
