---
title: 'Chapter 9 Practice: Data Dictionary'
author: "Tammy Lightle"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
---

## Load Data
```{r}
library(readr)
field_sightings <- read_csv("field_sightings_expanded.csv")

# inspect
head(field_sightings)
str(field_sightings)
```

# Create table for data dictionary
```{r}
# Create a new data frame for the data dictionary
data_dictionary <- data.frame(    
  
# Get the column names from the dataset, store in column 1 named "Field_Name"
  Field_Name = names(field_sightings),  
  
# Initialize the "Meaning" column with empty strings
  Meaning = rep("", ncol(field_sightings)),  

# Get the data class type of each column using class() and store in column 3 named "Type"
  Type = vapply(field_sightings, function(x) 
    paste(class(x), collapse = ", "), character(1)),    
 
# Initialize the "Units" column with empty strings 
  Units = rep("", ncol(field_sightings)),              

# Initialize the "Codes_Values" column with empty strings
  Codes_Values = rep("", ncol(field_sightings)), 

# Initialize the "Missing_Values" column with empty strings 
  Missing_Values = rep("", ncol(field_sightings)), 

# Initialize the "Valid_Values" column with empty strings
  Valid_Values = rep("", ncol(field_sightings)),    

# Initialize the "Notes_Relationships" column with empty strings
  Notes_Relationships = rep("", ncol(field_sightings)),

# Do not convert strings to factors (keep them as character vectors)
  stringsAsFactors = FALSE
)

# check
head(data_dictionary)
```

## Explore data
```{r}
table(field_sightings$Sex_Code, useNA = "ifany")
```
1,985 female (code = 1) and 1,801 male (code = 2), and 684 undetermined (left blank).

## Subassignment: Sex_Code
```{r}
row <- which(data_dictionary$Field_Name == "Sex_Code")
row
```
```{r}
data_dictionary[row, "Meaning"] <- "Sex of the bird when it could be determined visually."
data_dictionary[row, "Units"] <- "N/A"
data_dictionary[row, "Codes_Values"] <- "1 = female, 2 = male"
data_dictionary[row, "Missing_Values"] <- "NA means sex could not be determined"
data_dictionary[row, "Valid_Values"] <- "1, 2, or NA"
data_dictionary[row, "Notes_Relationship"] <- "Matches sex (f/m); more NA for wild hatch hatch-year birds."

tail(data_dictionary)
```

## Subassignment: Age_Code
```{r}
table(field_sightings$Age_Code, useNA = "ifany")
```
Age code (1) is hatch-year (n = 1,238), age code (2) is second-year (n = 1,094), age code (3) is adult (n = 1,742), and age code (99) is undertermined.

```{r}
row <- which(data_dictionary$Field_Name == "Age_Code")

data_dictionary[row, "Meaning"] <- "Age of the bird derived from known hatch year or visually estimated for wild hatch."
data_dictionary[row, "Units"] <- "N/A"
data_dictionary[row, "Codes_Values"] <- "1 = hatch-year, 2 = second-year, 3 = adult, 99 = unknown"
data_dictionary[row, "Missing_Values"] <- "NA and 99 mean age could not be determined"
data_dictionary[row, "Valid_Values"] <- "1, 2, 3, 99, or NA"
data_dictionary[row, "Notes_Relationship"] <- "Age code for wild hatch birds is estimated visually or unknown"
```

## Subassignment: Band_Color
```{r}
table(field_sightings$Band_Color, useNA = "ifany")
```
Band color encodes hatch year of birds from the reintroduction program. Silver is 2023 (n = 67), blue is 2024 (n = 138), green is 2025 (n = 97), and none indicates a wild-hatch bird (n = 4168).

```{r}
row <- which(data_dictionary$Field_Name == "Band_Code")

data_dictionary[row, "Meaning"] <- "Color of the bird's leg band."
data_dictionary[row, "Units"] <- "N/A"
data_dictionary[row, "Codes_Values"] <- "Silver = 2023, Blue = 2024, Green = 2025, None = no band"
data_dictionary[row, "Missing_Values"] <- "None means the bird does not have a leg band"
data_dictionary[row, "Valid_Values"] <- "Silver, Blue, Green, None"
data_dictionary[row, "Notes_Relationship"] <- "Band code for wild hatch birds is None"
```

## Subassignment: Site_Code
```{r}
table(field_sightings$Site_Code, useNA = "ifany")
```
820 records from basin rim (BR), 737 records from lakeside cove (LC), 871 records from meadow ranch (MR), 649 records from north pool (NP), 743 records from south riprarian (SR), and 650 records from willow channel (WC).

```{r}
row <- which(data_dictionary$Field_Name == "Site_Code")

data_dictionary[row, "Meaning"] <- "Wetland location."
data_dictionary[row, "Units"] <- "N/A"
data_dictionary[row, "Codes_Values"] <- "BR = Basin Rim, LC = Lakeside Cove, MR = Meadow Ranch, NP = North Pool, SR = South Riparian, WC = Willow Channel"
data_dictionary[row, "Missing_Values"] <- "N/A"
data_dictionary[row, "Valid_Values"] <- "BR, LC, MR, NP, SR, WC"
data_dictionary[row, "Notes_Relationship"] <- "N/A"
```


## Save file
```{r}
write.csv(data_dictionary, "field_sightings_dictionary.csv", row.names = FALSE)
```



