---
title: "Chapter 12 Practice"
author: "Tammy Lightle"
date: today
toc: true
toc-depth: 2
number-sections: true
format: 
  html:
    code-fold: true
    self-contained: true
execute:
  warning: true
  message: true
editor: visual
---

## Setup

Load libraries

```{r}
#| warning: false

library(dplyr)
library(dslabs)
```

## Task 1: Concept check

Vectorized conditional functions such as ifelse() and case_when() returns a new vector with one result for every position because each element is evaluated and returned, producing a vector of the same length as the original.

## Task 2: Simple ifelse()

Inspect dslabs::heights

```{r}
heights <- dslabs::heights
names(heights)
```

The heights data frame has two columns, sex and height.

Assign "Tall" to heights greater than 68, and "Not Tall" to all others.

```{r}
height_label <- ifelse(
  heights$height > 68,
  "Tall",
  "Not Tall"
)

table(height_label)
```

There are 542 entries with a height over 68, and 508 entries otherwise.

## Task 3: Recode with mutate() and case_when()

Assign mpg classes to mtcars dataset.

```{r}
mpg_recode <- mtcars %>%
  mutate(mpg_class = case_when(mpg > 25 ~ "High MPG",
                               mpg < 15 ~ "Low MPG",
                               TRUE ~ "Medium MPG"))

table(mpg_recode$mpg_class)
```

6 cars have mpg greater than 25 (high mpg), 21 cars have mpg between 15 and 25 (medium mpg), and 5 cars have mpg less than 15 (low mpg). Regardless, everyone is suffering with the increased gas prices.

## Task 4: Recode: name focus

Assign temperature group to each star based on their tempature.

```{r}
# inspect stars dataset
stars <- dslabs::stars
colnames(stars)
summary(stars$temp)

# code stars based on temp range
stars_recode <- stars %>%
  mutate(temp_group = case_when(temp < 4000 ~ "Cool",
                                temp <= 7500 ~ "warm",
                                TRUE ~ "Hot"))

table(stars_recode$temp_group)
```

There are 35 stars in the Cool category, 23 stars in the Warm category, and 38 stars in the Hot category.

## Task 5: Unit conversion with mutate()

Convert the height column from inches into cm.

```{r}
# inspect heights
head(heights)

# convert height in inches to cm
international_friendly_heights <- heights %>%
  mutate(height = height * 2.54)

head(international_friendly_heights)
```

The height column in `international_friendly_heights` is now in cm. The column replaced the original column that was in inches, therefore the data frame still has a total of 2 columns.
