---
title: "Chapter 15 Practice"
author: "Tammy Lightle"
date: today
toc: true
toc-depth: 2
number-sections: true
format: 
  html:
    code-fold: true
    self-contained: true
execute:
  warning: true
  message: true
editor: visual
---

## Task 1

Simulate 300 rolls of a loaded or fair 20-sided die and store in `big_die`.

```{r}
set.seed(200)

big_die <- sample(1:20, 300, TRUE)

any(big_die == 20)
```

Rolling a fair 20-sided die 300 times resulted in at least one roll where the value = 20.

## Task 2

**Die test**

Simulate 5,221 rolls of a fair 20-sided die. Create a table of how many times each face appeared.

```{r}
boring_weekend <- sample(1:20, 5221, TRUE)

d20_counts <- table(boring_weekend)

print(d20_counts)
```

What proportion of rolls were natural 20s?

```{r}
mean(boring_weekend == 20)
```

About `{r} round(mean(boring_weekend == 20) * 100, 1)`% of rolls were naturally 20.

**Baby blanket test**

Simulate 100 newborn lengths (inches) with mean 20.0, and sd 0.9. Check if there are any babies longer than 23 inches, and if so how many.

```{r}
blanketed_babies = rnorm(100, mean = 20.0, sd = 0.9)

any(blanketed_babies > 23)
sum(blanketed_babies > 23)
```

There are no babies that are longer than 23 inches. Based on this sample, the hospital does not need longer blankets.

Now, simulate for 1,000 weeks. For each week, determine how many babies were longer than 23 inches.

```{r}
weeks_too_long <- replicate(1000, rnorm(100, mean = 20.0, sd = 0.9))

any(weeks_too_long > 23)
sum(weeks_too_long > 23)
```

This was a 1,000 week simulation where each week had 100 babies with a mean length of 20.0 inches and a standard deviation of 0.9. There were `{r} sum(weeks_too_long > 23)` instances where a baby was longer than 23 inches. The hospital's need for longer blankets is slim, and represents only `{r} round(mean(weeks_too_long > 23) * 100, 2)`% of babies born.

```{r}
threshold <- 23
max_len <- max(weeks_too_long)
min_len <- min(weeks_too_long)
breaks_int <- seq(min_len - 1, max_len + 1, by = 0.5)

hist(weeks_too_long,
     breaks = breaks_int, 
     col = "grey",
     main = "Distribution of Baby Lengths (1000 reps of 100 babies)",
     xlab = "Baby Lengths (in)")

abline(v = threshold, col = "darkred", lwd = 2, lty = 2)
abline(v = mean(weeks_too_long), col = "darkblue", lwd = 2, lty = 1)

legend("topright",
       legend = c("Max Blanket Length (23)", "Mean Length"),
       col = c("darkred", "darkblue"),
       lty = c(2, 1),
       lwd = 2,
       bg = "white",
       bty = "o")
```
