---
title: "Signature Assignment"
subtitle: "Descriptive Statistics"
author: "Tammy Lightle"
date: today
toc: true
toc-depth: 2
number-sections: true
format: 
  html:
    code-fold: true
    self-contained: true
execute:
  warning: true
  message: true
editor: visual
---

# Part 1

## Load packages and data

```{r}
#| warning: false

library(dslabs)
library(dplyr)
library(palmerpenguins)
data("penguins")
```

The data have three penguin species observed in the Palmer Archipelago, Antartica, collected by Dr. Kristen Gorman with Palmer Station LTER.

**Research question: How do penguin body measurements differ by species, sex, and island in the Palmer Archipelago?**

## Data type

What are the column names in this dataset?

The column names in the dataset `penguins` are `{r} colnames(penguins)`.

The `species` column is stored as `{r} class(penguins$species)` in R. It has the unique values: `{r} unique(penguins$species)`.

The `island` column is stored as `{r} class(penguins$island)` in R. It has the unique values: `{r} unique(penguins$island)`.

The `bill_length_mm` column is stored as `{r} class(penguins$bill_length_mm)` in R. The data are continuous.

The `bill_depth_mm` column is stored as `{r} class(penguins$bill_depth_mm)` in R. The data are continuous.

The `flipper_length_mm` column is stored as `{r} class(penguins$flipper_length_mm)` in R. The data are continuous.

The `body_mass_g` column is stored as `{r} class(penguins$body_mass_g)` in R. The data are continuous.

The `sex` column is stored as `{r} class(penguins$sex)` in R. It has the unique values: `{r} unique(penguins$sex)`.

The `year` column is stored as `{r} class(penguins$year)` in R.

This dataset is a sample because it represents a subset of the penguin population which was observed, not a count or measurement of the entire penguin population. Species, island, and sex are categorical whereas bill length, bill depth, flipper length, and body mass are numerical. It would make sense to count the categorical variables, and summarize the numerical variables with mean, median, and spread. Year is stored as numeric, but should be interpreted carefully before averaging.

# Part 2

## Task 1: Count-based summaries

**Step 1: choose appropriate variables**

1.  I could use length() on any column in order to figure out how many rows are in the column.
2.  I could use count() on the categorical columns to figure out the frequency of each category. These columns are `species`, `island`, and `sex`.
3.  Mode() could be useful on any column, depending on the question that is trying to be answered. Let's choose `year` to learn which year has the most observations present.

**Step 2: Length report**

The `species` column in the `penguins` dataset contains `{r} length(penguins$species)` observations.

I chose `species` for length() because it is a `{r} class(penguins$species)` variable. It makes sense to use length() because I want to know how many species observations exist in the dataset.

**Step 3: Frequency table**

Create frequency table for `island`.

```{r}
island_counts <- penguins %>%
  count(island)

island_counts
```

There were 168 observations on Biscoe, 124 observations on Dream, and 52 observations on Torgersen.

**Step 4: Identify the mode**

The mode of `island` is Biscoe, which makes up `{r} penguins %>% filter(island == "Biscoe") %>% count()` out of `{r} length(penguins$island)` total observations. I chose `island` for count() because it is a `{r} class(penguins$island)` variable. It makes sense to use count because I wanted to know the frequency of each category.

## Task 2: Calculate median

Calculate median of a column in `penguins` dataset.

```{r}
median(penguins$bill_length_mm, na.rm = TRUE)
```

The median of the `bill_length_mm` column in the `penguins` dataset is `{r} median(penguins$bill_length_mm, na.rm = TRUE)`.

I chose the `bill_length_mm` variable from the `penguins` dataset to describe with a median because it is a `{r} class(penguins$bill_length_mm)` variable that may contain outliers. The median is a better measure of center for this variable because it is not affected by extreme values, unlike the mean. Using the median allows us to get a more accurate representation of the typical value in the dataset for this variable.

## Task 3: Calculate mean

Calculate the mean of a column in the `penguins` dataset. I'll choose the `bill_length_mm` column.

```{r}
mean(penguins$bill_length_mm, na.rm = TRUE)
```

The mean of the `bill_length_mm` column in the `penguins` dataset is `{r} mean(penguins$bill_length_mm, na.rm = TRUE)`.

I chose the `bill_length_mm` variable from the `penguins` dataset to describe with a median because it is a `{r} class(penguins$bill_length_mm)` variable. The mean is a good measure of center for this variable because it takes into account all values in the dataset and provides an average that represents the typical value.

## Task 4: Compare measures of center {#sec-center}

For the `bill_length_mm` column, the median is `{r} median(penguins$bill_length_mm, na.rm = TRUE)` and the mean is `{r} mean(penguins$bill_length_mm, na.rm = TRUE)`. The mean and the median are close together, however not exact. The mean is less than the median, indicating a negatively skewed distribution, though not much.

There are not unusually high or low values, so the mean would be the better measure to report for this variable because it represents the full dataset. If, however, you had a variable with high or low outliers, the median would be the better measure to report because it is not affected by those outliers.

## Task 5: Compare measures of center

I'll chose `bill_length_mm` as my numerical variable and `island` as my categorical grouping variable to compare bill length by island. Maybe the food source at each island is different, and therefore the penguins have evolved with different bill lengths.

Since we learned in Task 4 that there are not outliers in the `bill_length_mm` variable, I'll calculate the mean for each group.

```{r}
mean(penguins$bill_length_mm[penguins$island == "Torgersen"], na.rm = TRUE)
mean(penguins$bill_length_mm[penguins$island == "Biscoe"], na.rm = TRUE)
mean(penguins$bill_length_mm[penguins$island == "Dream"], na.rm = TRUE)
```

The mean bill length for penguins observed on Torgersen is `{r} mean(penguins$bill_length_mm[penguins$island == "Torgersen"], na.rm = TRUE)` mm, while the mean bill length for penguins observed on Biscoe is `{r} mean(penguins$bill_length_mm[penguins$island == "Biscoe"], na.rm = TRUE)` mm, and the mean bill length for penguins observed on Dream is `{r} mean(penguins$bill_length_mm[penguins$island == "Dream"], na.rm = TRUE)` mm.

There is a difference in the mean bill length among penguins observed on different islands, however, more research will need to be done to determine if this is of significance. Early observations show that the average bill length of penguins observed on Biscoe is about `{r} round(mean(penguins$bill_length_mm[penguins$island == "Biscoe"], na.rm = TRUE) - mean(penguins$bill_length_mm[penguins$island == "Torgersen"], na.rm = TRUE), 2)` mm greater than those observed on Torgersen.

Shown in Task 1, Torgersen had much fewer observations than both Biscoe and Dream. This difference in number of observations may help explain the difference in mean bill lengths. Solely looking at the mean is not enough to tell if the island has a significant effect on the penguin's bill length.

# Part 3

## Standard Deviation {#sec-sd}

The standard deviation calculation is as follows: $$
s = \sqrt{s^2} = \sqrt{\frac{\sum_{i=1}^{n}(x_i - \bar{x})^2}{n-1}}
$$

Calculate the standard deviation for`body_mass_g`.

```{r}
sd(penguins$body_mass_g, na.rm = TRUE)
```

The standard deviation of the `body_mass_g` column in the `penguins` dataset is `{r} round(sd(penguins$body_mass_g, na.rm = TRUE), 2)`.

::: callout-note
I chose the `body_mass_g` variable because it is a numerical measurement. The standard deviation tells me how spread out the values are around the mean. A larger standard deviation means the values vary more, while a smaller standard deviation means the values are more tightly clustered.
:::

::: callout-note
Compared with the center measures discussed in @sec-center, the standard deviation helps describe how much the values vary around the mean.
:::

## Range {#sec-range}

Use `range()` to find the minimum and maximum of `body_mass_g`.

```{r}
penguin_range <- range(penguins$body_mass_g, na.rm = TRUE)

penguin_range
```

```{r}
penguin_range_value <- penguin_range[2] - penguin_range[1]
```

-   **Minimum:** `{r} range(penguins$body_mass_g, na.rm = TRUE)[1]`

-   **Maximum:** `{r} range(penguins$body_mass_g, na.rm = TRUE)[2]`

-   **Range:** `{r} diff(range(penguins$body_mass_g, na.rm = TRUE))`

## Interquartile Range {#sec-iqr}

Calculate the interquartile range of `body_mass_g`.

```{r}
iqr_value <- IQR(penguins$body_mass_g, na.rm = TRUE)
```

The interquartile range of the `body_mass_g` column in the `penguins` dataset is `{r} iqr_value`.

The IQR describes the spread of the middle 50% of the data. It is useful because it is less affected by unusually high or low values than the full range. This makes it a strong choice when I want to describe the spread of the typical values in the dataset.

## Five-Number Summary {#sec-five}

Use `summary()` to get an overview.

```{r}
summary(penguins$body_mass_g)
```

In the `body_mass_g` column, the minimum value is `{r} summary(penguins$body_mass_g)[1]`, the median is `{r} summary(penguins$body_mass_g)[3]`, and the maximum is `{r} summary(penguins$body_mass_g)[5]`. The first and third quartiles show where the middle half of the data falls. This summary gives me a quick picture of both center and spread in this variable.

## Early Takeaways

For the `body_mass_g` variable, I think the most useful measure of spread is the IQR because it unaffected by the min and max values in the dataset.

# Part 4

## Describing data with `dplyr`
