---
title: "Wildfire Signature Assignment"
author: "Tammy Lightle"
date: today
toc: true
toc-depth: 2
number-sections: true
format: 
  html:
    code-fold: true
    self-contained: true
execute:
  warning: true
  message: true
editor: visual
---

## Load libraries and data

Load the U.S. Forest Service's Wildfire Hazard Potential (WHP) dataset. The WHP is a unitless, continuous index, where a greater value is equivalent to a greater relative hazard.

```{r}
#| warning: false

library(tidyverse)

county_raw <- read.csv("whp_raw_sample.csv")

colnames(county_raw)
nrow(county_raw)
head(county_raw)
```

The data frame has `{r} nrow(county_raw)` rows of data, and four columns: `{r} colnames(county_raw)`. Each row represents a raster cell with an associated wildfire hazard potential (`whp`). Because multiple raster cells can fall within a county, each county appears in multiple rows. The variables `county` and `county_fips` identify the county for each raster cell, and `cell_area_km2` represents the area (in square kilometers) of each cell.

## Basic Summary

Summarize the total number of raster cells (`n_rows`), the number of missing values in `whp`, and the minimum and maximum values of raster cell area.

```{r}
county_summary <- county_raw %>%
  summarize(n_rows = n(),
            n_wa_whp = sum(is.na(whp)),
            min_area = min(cell_area_km2),
            max_area = max(cell_area_km2))

county_summary
```
