---
title: "Wildfire Signature Assignment"
author: "Tammy Lightle"
date: today
toc: true
toc-depth: 2
number-sections: true
format: 
  html:
    code-fold: true
    self-contained: true
execute:
  warning: true
  message: true
editor: visual
---

# Part A

## Load libraries and data

Load the U.S. Forest Service's Wildfire Hazard Potential (WHP) dataset. The WHP is a unitless, continuous index, where a greater value is equivalent to a greater relative hazard.

```{r}
#| warning: false

library(tidyverse)

county_raw <- read.csv("whp_raw_sample.csv")

colnames(county_raw)
nrow(county_raw)
head(county_raw)
```

The data frame has `{r} nrow(county_raw)` rows of data, and four columns: `{r} colnames(county_raw)`. Each row represents a raster cell with an associated wildfire hazard potential (`whp`). Because multiple raster cells can fall within a county, each county appears in multiple rows. The variables `county` and `county_fips` identify the county for each raster cell, and `cell_area_km2` represents the area (km^2^) of each cell.

## Basic Summary

Summarize the total number of raster cells (`n_rows`), the number of missing values in `whp`, and the minimum and maximum values of raster cell area.

```{r}
county_summary <- county_raw %>%
  summarize(n_rows = n(),
            n_na_whp = sum(is.na(whp)),
            min_area = min(cell_area_km2),
            max_area = max(cell_area_km2))

county_summary
```

There are no NA values in `whp`, indicating that each row has a valid value. The range of raster cell area is only `{r} county_summary$max_area - county_summary$min_area` km^2^, and no cells have an area of 0 km^2^ which is expected.

## Compute WHP Quantile Edges

Calculate the quantiles of the `whp` column. Use probabilities 0, 0.25, 0.50, 0.75, 0.90, 0.95, 0.99, 1.00.

```{r}
e <- quantile(county_raw$whp,
              probs = c(0.0, .25, .5, .75, .9, .95, .99, 1.0), 
              na.rm = TRUE)

e
```

`e` is a vector containing the percentiles of the `whp` field.

## Assign a Categorical WHP Risk Label

Assign a risk label based on the WHP percentile. The risk label should be an ordered factor following the levels Very Low, Low, Moderate, High, Very High, Extreme, and Ultra (Top 1%).

```{r}
county_cats <- county_raw %>%
  mutate(category = case_when(
    whp >= e[1] & whp < e[2] ~ "Very Low",
    whp >= e[2] & whp < e[3] ~ "Low",
    whp >= e[3] & whp < e[4] ~ "Moderate",
    whp >= e[4] & whp < e[5] ~ "High",
    whp >= e[5] & whp < e[6] ~ "Very High",
    whp >= e[6] & whp < e[7] ~ "Extreme",
    whp >= e[7] & whp <= e[8] ~ "Ultra (Top 1%)",
    .default = NA))

#check
county_cats %>%
  group_by(category) %>%
  summarize(cat_count = n()) %>%
  arrange(cat_count)
```

The distribution of whp values is skewed toward the low side of categories.

## Calculate Random Ignition Probability

Calculate the total raster cell area in each category, then determine the probability of random ignition based on the proportion of area in each category.

```{r}
prob_tbl <- county_cats %>%
  group_by(category) %>%
  summarize(area_km2 = sum(cell_area_km2),
            n_cells = n(),
            .groups = "drop") %>%
  mutate(probability = area_km2 / sum(area_km2)) %>%
  select(category, probability, area_km2, n_cells)

# check
sum(prob_tbl$probability)
prob_tbl %>% arrange(probability)
```

`prob_tbl` shows the probability that a random ignition would occur within each risk category. This is assuming that the ignition is equally likely anywhere in the state.

For example, the nearly 4% probability for the `Extreme` risk category means about 4% of Idaho's land area falls into that category. In other words, a random ignition would have a 4% chance of landing in a raster cell whose WHP value is in the extreme (95-99th percentile) range.
