---
title: "Wildfire Signature Assignment"
author: "Tammy Lightle"
date: today
toc: true
toc-depth: 2
number-sections: true
format: 
  html:
    code-fold: true
    self-contained: true
    theme: cosmo
execute:
  warning: true
  message: true
editor: visual
---

```{r setup}
#| warning: false

library(tidyverse)

county_raw <- read.csv("whp_raw_sample.csv")

oxford_comma <- function(x) {
  if(length(x) == 0) return("")
  if(length(x) == 1) return(x)
  if(length(x) == 2) return(paste(x, collapse = "and "))
  paste0(paste(x[-length(x)], collapse = ", "), ", and ", x[length(x)])
}
```

# Part A

## Introduction

This analysis uses the U.S. Forest Service's Wildfire Hazard Potential (WHP) dataset. WHP is a unitless, continuous index, where higher values indicate greater relative hazard. Each record represents a raster cell, including the county it belongs to, the county identifier, and the WHP value for that cell. Since multiple raster cells can fall within a county, each county appears in multiple rows.

```{r}
#| results: 'hide'

colnames(county_raw)
nrow(county_raw)
head(county_raw)
```

The dataset contains four columns: `{r} oxford_comma(colnames(county_raw))`.

-   `whp` is `{r} class(county_raw$whp)`, and ranges from `{r} min(county_raw$whp)` to `{r} max(county_raw$whp)`.

-   `county` is `{r} class(county_raw$county)`.

-   `county_fips` is `{r} class(county_raw$county_fips)`, but acts as an identifier and is not intended to be used in calculations.

-   `cell_area_km2` is `{r} class(county_raw$cell_area_km2)`, and ranges from `{r} min(county_raw$cell_area_km2)` km^2^ to `{r} max(county_raw$cell_area_km2)` km^2^ .

```{r}
#| results: 'hide'

county_summary <- county_raw %>%
  summarize(n_rows = n(),
            n_na_whp = sum(is.na(whp)),
            n_na_area = sum(is.na(cell_area_km2)),
            min_area = min(cell_area_km2),
            max_area = max(cell_area_km2))

county_summary
```

The dataset has `{r} nrow(county_raw)` rows of data, and there are no NA values in `whp`, indicating that each row has a valid value. The range of raster cell area is very small (`{r} county_summary$max_area - county_summary$min_area` km^2^), and no cells have an area of 0 km^2^, as expected.

## Compute WHP Quantile Edges

The quantiles of the `whp` column are calculated using the probabilities: 0, 0.25, 0.50, 0.75, 0.90, 0.95, 0.99, 1.00.

```{r}
e <- quantile(county_raw$whp,
              probs = c(0.0, .25, .5, .75, .9, .95, .99, 1.0), 
              na.rm = TRUE)

e
```

`e` is a vector containing the percentiles of the `whp` field.

## Assign a Categorical WHP Risk Label

A risk label is assigned to each raster cell based on the WHP percentile. The risk label is an ordered factor following the levels Very Low, Low, Moderate, High, Very High, Extreme, and Ultra (Top 1%).

```{r}
county_cats <- county_raw %>%
  mutate(category = case_when(
    whp >= e[1] & whp < e[2] ~ "Very Low",
    whp >= e[2] & whp < e[3] ~ "Low",
    whp >= e[3] & whp < e[4] ~ "Moderate",
    whp >= e[4] & whp < e[5] ~ "High",
    whp >= e[5] & whp < e[6] ~ "Very High",
    whp >= e[6] & whp < e[7] ~ "Extreme",
    whp >= e[7] & whp <= e[8] ~ "Ultra (Top 1%)",
    .default = NA))

#check
county_cats %>%
  group_by(category) %>%
  summarize(cat_count = n()) %>%
  arrange(cat_count)
```

The distribution of `whp` values is skewed toward the lower categories.

## Calculate Random Ignition Probability

The probability of random ignition is determined based on the proportion of area in each category.

```{r}
#| label: tbl-probability
#| tbl-cap: "Random Ignition Probability by Risk Category"

prob_tbl <- county_cats %>%
  group_by(category) %>%
  summarize(area_km2 = sum(cell_area_km2),
            n_cells = n(),
            .groups = "drop") %>%
  mutate(probability = area_km2 / sum(area_km2)) %>%
  select(category, probability, area_km2, n_cells) %>%
  arrange(probability)

knitr:: kable(prob_tbl, digits = 3)
```

```{r}
#check
sum(prob_tbl$probability)
```

@tbl-probability shows the probability that a random ignition would occur within each risk category. This assumes that a random ignition is equally likely anywhere in the state.

For example, the 4% probability for the `Extreme` risk category means about 4% of Idaho's land area falls into that category. In other words, a randomly occuring ignition has a 4% chance of landing in a raster cell with a WHP value in the Extreme (95-99th percentile) range.

# Part B
