---
title: "Wildfire Signature Assignment"
author: "Tammy Lightle"
date: today
toc: true
toc-depth: 2
number-sections: true
format: 
  html:
    code-fold: true
    self-contained: true
    theme: cosmo
execute:
  warning: true
  message: true
editor: visual
---

```{r setup}
#| warning: false

library(tidyverse)

county_raw <- read.csv("whp_raw_sample.csv")

oxford_comma <- function(x) {
  if(length(x) == 0) return("")
  if(length(x) == 1) return(x)
  if(length(x) == 2) return(paste(x, collapse = "and "))
  paste0(paste(x[-length(x)], collapse = ", "), ", and ", x[length(x)])
}
```

# Part A

## Dataset Description {#sec-dataset-description}

This analysis uses the U.S. Forest Service's Wildfire Hazard Potential (WHP) dataset. WHP is a unitless, continuous index where higher values indicate greater relative wildfire hazard. Each record represents a raster cell and includes the county name, county identifier, and the WHP value for that cell. Since multiple raster cells can fall within a county, each county appears in multiple rows.

```{r}
#| results: 'hide'

colnames(county_raw)
nrow(county_raw)
head(county_raw)
```

The dataset contains four columns: `{r} oxford_comma(colnames(county_raw))`.

-   `whp` is `{r} class(county_raw$whp)`, and ranges from `{r} min(county_raw$whp)` to `{r} max(county_raw$whp)`.

-   `county` is `{r} class(county_raw$county)`.

-   `county_fips` is `{r} class(county_raw$county_fips)`, but acts as an identifier and is not intended to be used in calculations.

-   `cell_area_km2` is `{r} class(county_raw$cell_area_km2)`, and ranges from `{r} min(county_raw$cell_area_km2)` km^2^ to `{r} max(county_raw$cell_area_km2)` km^2^ .

```{r}
#| results: 'hide'

county_summary <- county_raw %>%
  summarize(n_rows = n(),
            n_na_whp = sum(is.na(whp)),
            n_na_area = sum(is.na(cell_area_km2)),
            min_area = min(cell_area_km2),
            max_area = max(cell_area_km2))

county_summary
```

The dataset has `{r} nrow(county_raw)` rows of data, and there are no NA values in `whp`, indicating that each row has a valid value. The range of raster cell area is very small (`{r} county_summary$max_area - county_summary$min_area` km^2^), and no cells have an area of 0 km^2^, as expected.

## Quantiles {#sec-quantiles}

The quantiles of the `whp` column are calculated using the probabilities: 0, 0.25, 0.50, 0.75, 0.90, 0.95, 0.99, 1.00.

```{r}
e <- quantile(county_raw$whp,
              probs = c(0.0, .25, .5, .75, .9, .95, .99, 1.0), 
              na.rm = TRUE)

e
```

`e` is a numeric vector containing the selected quantiles of the `whp` variable.

## Risk Categories {#sec-risk-categories}

A risk label is assigned to each raster cell based on the WHP percentile. The risk label is an ordered factor following the levels Very Low, Low, Moderate, High, Very High, Extreme, and Ultra (Top 1%).

```{r}
county_cats <- county_raw %>%
  mutate(category = case_when(
    whp >= e[1] & whp < e[2] ~ "Very Low",
    whp >= e[2] & whp < e[3] ~ "Low",
    whp >= e[3] & whp < e[4] ~ "Moderate",
    whp >= e[4] & whp < e[5] ~ "High",
    whp >= e[5] & whp < e[6] ~ "Very High",
    whp >= e[6] & whp < e[7] ~ "Extreme",
    whp >= e[7] & whp <= e[8] ~ "Ultra (Top 1%)",
    .default = NA))

#check
county_cats %>%
  group_by(category) %>%
  summarize(cat_count = n()) %>%
  arrange(cat_count)
```

The distribution of `whp` values is skewed toward the lower categories (right-skewed).

## Random Ignition Probability {#sec-random-ignition-probability}

The probability of random ignition is determined based on the proportion of area in each category.

```{r}
#| label: tbl-probability
#| tbl-cap: "Random Ignition Probability by Risk Category"

prob_tbl <- county_cats %>%
  group_by(category) %>%
  summarize(area_km2 = sum(cell_area_km2),
            n_cells = n(),
            .groups = "drop") %>%
  mutate(probability = area_km2 / sum(area_km2)) %>%
  select(category, probability, area_km2, n_cells) %>%
  arrange(probability)

knitr:: kable(prob_tbl, digits = 3)
```

```{r}
#check
sum(prob_tbl$probability)
```

@tbl-probability shows the probability that a random ignition would occur within each risk category. This assumes that a random ignition is equally likely to occur anywhere in the state.

For example, the 4% probability for the `Extreme` risk category means about 4% of Idaho's land area falls into that category. In other words, a randomly occurring ignition has a 4% chance of landing in a raster cell with a WHP value in the Extreme (95-99th percentile) risk category.

# Part B

## Probability of Spread {#sec-probability-of-spread}

Find the probability that the 8-hour straight-line spread is at least 10 km if the ignition starts in a `High` hazard area.

```{r}
p_high_exact <- plnorm(10, meanlog = 1.907, sdlog = 0.280, lower.tail = FALSE)

p_high_exact
p_high_exact <- p_high_exact * 100
p_high_exact
```

Using a lognormal model, there is a `{r} round(p_high_exact, 2)`% probability that an ignition in a `High` hazard area will spread at least 10 km over an 8-hour period.

## 95th Percentile Run Length {#sec-95th-percentile-run-length}

For ignitions in `High` hazard areas, find the 95th percentile run length for an 8-hour straight line spread.

```{r}
q95_high <- qlnorm(.95, meanlog = 1.907, sdlog = 0.280)

q95_high
```

Approximately 95% of ignitions in a `High` hazard area will spread no more than `{r} round(q95_high, 2)` km over 8 hours.

## Density {#sec-density}

Using a probablity density function, compare the relative likelihood of a spread distance of 3 km and 10 km.

```{r}
d_3 <- dlnorm(3, meanlog = 1.907, sdlog = 0.280)

d_10 <- dlnorm(10, meanlog = 1.907, sdlog = 0.280)

d_3
d_10
```

The density value for 3 km is `{r} round(d_3, 3)` and the density value for 10 km is `{r} round(d_10, 3)`. The higher density at 10 km suggests that distances near 10km will occur more frequently than those near 3 km.

## Simulate High-Hazard Distances {#sec-simulate-high-hazard-distances}

Using 450 simulations, find the range and typical values of 8-hour straight-line distances for ignitions in `High` hazard areas.

```{r}
set.seed(2025)

high_lognorm <- rlnorm(450, meanlog = 1.907, sdlog = 0.280)

# check
summary(high_lognorm)
length(high_lognorm)

range(high_lognorm)
```

Based on 450 simulated distances, median is `{r} round(median(high_lognorm), 2)` and the mean is `{r} round(mean(high_lognorm), 2)`, indicating that the data is right-skewed. The range of distances spans from `{r} round(min(high_lognorm), 2)` km to `{r} round(max(high_lognorm), 2)` km.

## Fast-Growing Ignitions {#sec-fast-growing-ignitions}

For this part of the analysis, a fast-growing ignition is one where the 8-hour straight-line distance is ≥ 10 km. Determine how many of the 450 simulated distances would be fast-growing.

```{r}
fast_growing_count <- (high_lognorm >= 10) %>%
  sum()

fast_growing_count
```

Out of 450 simulations, `{r} fast_growing_count` are ignitions that are fast-growing. The simulated proportion of fast-growing ignitions is `{r} round(fast_growing_count * 100 / length(high_lognorm), 2)`%, which is slightly lower than the probability calculated using the lognormal model (`{r} round(p_high_exact, 2)`%).

## Visualize Simulated Distances {#sec-visualize-simulated-distances}

A histogram is used to visualize the simulated distances.

```{r}
#| label: fig-sim-dist
#| fig-cap: "Simulated 8-hour straight-line distances for High hazard areas."
#| fig-alt: "A histogram showing the straight-line distances for High hazard areas with distance on the x-axis and frequency on the y-axis."

hist(high_lognorm,
     main = "Distribution of Simulated Distances (n = 450)",
     xlab = "Distance (km)",
     col = "steelblue")
```

@fig-sim-dist shows the distribution of simulated distances. The shape is right-skewed, which agrees with what was found in @sec-simulate-high-hazard-distances.

# Part C

## Create vectors

**Category Labels**

Create a vector of category labels.

```{r}
labels <- c("Very Low", "Low", "Moderate", "High", "Very High", "Extreme", "Ultra (Top 1%)")

labels
```

The character vector, `labels`, will be used to assign the order of risk categories, and must be spelled correctly to align with the risk categories assigned to the dataset in Part A.

**Probabilities**

Create a vector of probabilities.

```{r}
prob <- c(.25, .25, .25, .15, .05, .04, .01)

sum(prob)
```

The sum of the probabilities in `prob` equals 1, as expected.

**Mean logs**

Create a vector of the Mean log values.

```{r}
meanlog <- c(0.582, 1.046, 1.451, 1.907, 2.455, 2.866, 3.310)

meanlog
```

**SD logs**

Create a vector of the SD log values.

```{r}
sdlog <- c(0.472, 0.325, 0.325, 0.280, 0.246, 0.220, 0.212)

sdlog
```

**Check**

Ensure that the lengths of all vectors is equal so that each category has a value for probability, mean log, and sd log.

```{r}
length(labels) == length(prob) &&
length(labels) == length(meanlog) &&
length(labels) == length(sdlog)
```

## Simulate One Season

### Categories

Simulate one season of 450 ignitions to find the categories according to their probabilities. Display results in a table.

```{r}
set.seed(2025)

idx <- sample.int(length(labels),
              size = 450,
              replace = TRUE,
              prob = prob)

idx_tbl <- table(factor(labels[idx], levels = labels))

idx_tbl
```

Check the simulated category counts compared to the theoretical proportions in `prob`.

```{r}
(table(labels[idx]) / 450)[labels]

prob

```

That result is a bit hard to read, I want to visualize in a table instead.

```{r}
#| label: tbl-summary
#| tbl-cap: "Simulated vs Expected Probability"

summary_tbl <- data.frame(
  Category = labels,
  Frequency = as.numeric(idx_tbl),
  Sim_Prob = as.numeric(idx_tbl) / 450,
  Exp_Prob = prob
)

knitr:: kable(summary_tbl, digits = 2)
```

I used AI (chatGPT) to help organize this into a table so it is easier to compare.

@tbl-summary shows the risk category, frequency, simulated probability, and expected probability in the one-season simulation. The simulated probabilities closely approximate the expected probabilities, with small deviations at the high-risk (lower frequency) categories due to random sampling variability.

### Distances

Simulate 8-hour distances for the same season.

```{r}
d <- rlnorm(length(idx), meanlog = meanlog[idx], sdlog = sdlog[idx])

head(d)
summary(d)
```

The median distance for this simulation is `{r} round(median(d), 2)` and the mean is `{r} round(mean(d), 2)` indicating the data is right-skewed.

### Fast-Growing ignitions

A fast-growing ignition is one where distance ≥ 10 km, so create logical vector from `d` indicating if each value is above the fast-growing threshold.

```{r}
fast_flags_one <- d >= 10
```

Count the number of fast-growing ignitions in this simulation.

```{r}
fast_count_one <- sum(fast_flags_one)
```

In this simulation, there are `{r} fast_count_one` ignitions that are considered fast-growing, which represents `{r} round(mean(fast_flags_one) * 100, 1)` % of total ignitions.

### Combine Logic

Combine the steps into one code block so it can be used to replicate many seasons.

```{r}
set.seed(2025)

fast_count_one <- {
  idx <- sample.int(length(labels),
              size = 450,
              replace = TRUE,
              prob = prob)
  
  sum(rlnorm(length(idx), meanlog = meanlog[idx], sdlog = sdlog[idx]) >= 10)
}

fast_count_one
```

I was initially getting different results for `fast_count_one`, and used help from AI (chatGPT). I learned that I need to set the same seed for both versions so that both versions use the same random inputs. Now, I can see that the logic in this combined code block matches the logic created iteratively.

## Simulate Many Seasons

Simulate 1000 seasons, where each season has 450 ignitions.

```{r}
set.seed(2025)

counts_fast_mix <- replicate(1000,
          {idx <- sample.int(length(labels),
              size = 450,
              replace = TRUE,
              prob = prob)
          sum(rlnorm(length(idx), meanlog = meanlog[idx], sdlog = sdlog[idx]) >= 10)
          }
          )

summary(counts_fast_mix)
```

Find the number of seasons that have at least 50 fast-growing ignitions and the probability that any one season has at least 50 fast-growing ignitions.

```{r}
sum(counts_fast_mix >= 50)

prob_50 <- sum(counts_fast_mix >= 50) / length(counts_fast_mix)
```

In this simulation of 1000 seasons, there are `{r} sum(counts_fast_mix >= 50)` seasons that have at least 50 fast-growing ignitions. Therefore the probability that any one season has at least 50 fast-growing events is `{r} prob_50 * 100`%.

### Distribution of Fast-Growing Counts

The distribution of fast-growing counts can be visualized using a histogram, where the x-axis represents the total number of fast-growing counts in a season.

```{r}
#| label: fig-fast-counts
#| fig-cap: ""
#| fig-alt: "A histogram showing the number of fast-growing events per season, with the number of fast-growing events on the x-axis and frequency on the y-axis."

hist(counts_fast_mix,
     main = "Distribution of Seasonal Fast-Growing Ignitions",
     xlab = "Number of Fast-Growing Ignitions",
     col = "steelblue")

abline(v = 50, col = "coral2", lwd = 2, lty = 1)

legend("topright",
       legend = "Fast-Fire Threshold",
       col = "coral2",
       lty = 1,
       lwd = 2,
       bg = "white",
       bty = "o")
```

@fig-fast-counts shows the distribution of seasonal fast-growing ignitions. The number of seasons that visually appear above the threshold of at least 50 fast-growing ignitions appears to match the calculated probability of `{r} round(prob_50 * 100, 1)`%.
